---
title: "MDA Milestone 1"
output: github_document
date: "2024-10-03"
---

# Welcome to your (maybe) first-ever data analysis project!
  
  And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2024> Organization. You can do this by following the steps found on canvas in the entry called [MDA: Create a repository](https://canvas.ubc.ca/courses/158528/pages/mda-create-a-repository). One completed, your repository should automatically be listed as part of the stat545ubc-2024 Organization.

# Instructions

## For Both Milestones

-   Each milestone has explicit tasks. Tasks that are more challenging will often be allocated more points.

-  Each milestone will be also graded for reproducibility, cleanliness, and coherence of the overall Github submission. 

- While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R.html) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

- The milestones will be equally weighted. 

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 36 points: 30 for your analysis, and 6 for overall reproducibility, cleanliness, and coherence of the Github submission. 

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset 

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**
  
  -   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!
  
  1.1 **(1 point)** Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:
  
  **Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1. cancer_sample

  The cancer sample dataset has information on patients with benign and malignant tumors and also information on the size, shape, area, and texture of each of the tumors. 

2. vancouver_trees

  The vancouver trees dataset has information on different species of trees planted in vancouver. There is information on where the trees are planted as well as taxonomy and size. 

3. apt_buildings

  The apartment buildings dataset contains information on the features in each of the apartments and if the utilities are included in the apartment or if they are separate. 

4. building_permits

  The building permits dataset shows data collected for building projects for different buildings in Vancouver and contains information on where the building is and what type of work will be done.

<!----------------------------------------------------------------------------->

1.2 **(6 points)** One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->
##### Dataset 1: cancer_sample
```{r}
cancer_sample %>%
group_by(diagnosis) %>%
summarise(count = n())

cancer_sample %>%
summarize(num_rows = n())

glimpse(cancer_sample)
```
##### Dataset 2: vancouver_trees
```{r}
vancouver_trees %>%
group_by(species_name) %>%
summarise(count = n())

vancouver_trees %>%
summarize(num_rows = n())

glimpse(vancouver_trees)
```
##### Dataset 3: apt_buildings
```{r}
apt_buildings %>%
group_by(year_built) %>%
summarise(count = n())

apt_buildings %>%
summarise(num_rows = n())

glimpse(apt_buildings)
```
##### Dataset 4: building_permits
```{r}
building_permits %>%
group_by(property_use) %>%
summarise(count = n())

building_permits %>%
summarise(num_rows = n())

glimpse(building_permits)
```
<!----------------------------------------------------------------------------->


1.3 **(1 point)** Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 1. What lead you to choose this one? Briefly explain your choice below.

<!-------------------------- Start your work below ---------------------------->
I am going to use the vancouver_trees dataset. I chose this dataset because after inspecting the data, it includes information on where the trees are located as well as the species of tree and numerical measurements on the size of the tree, measured by height and diameter. I think this dataset is more interesting and will be easier to work with, compared to the cancer sample dataset, which I originally wanted to work with. The cancer sample dataset only includes numeric information on the size and shape of the tumors, not where in the body the cancer was or what type of cancer it was,  which I would have found more interesting.  
<!----------------------------------------------------------------------------->

1.4 **(2 points)** Time for a  final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with your dataset. Note it down below.

<!-------------------------- Start your work below ---------------------------->
Is there a relationship between when the trees were planted and how tall the trees are now, or is the height of the tree more strongly correlated to the species of tree?
<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset 

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 **(12 points)** Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. 

Make sure that you're using dplyr and ggplot2 rather than base R for this task. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 **(4 points)** For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->
1. Plot distribution of numeric variable
```{r}
ggplot(vancouver_trees, aes(diameter)) +
  geom_histogram(aes(fill = diameter), fill = "#D55E00", color = "black", binwidth = 5) +
  labs(title = "Distribution of Tree Diameter", x = "Diameter", y = "Count")+
  xlim(0, 435) + 
  theme_minimal()

```

I chose to plot the numeric variable diameter in the vancouver trees dataset. I represented this as a bar graph to show how many trees have the same diameter. We can see that majority of the trees have a diameter of less than 50 (Note: units not provided in the dataset). 10 trees have a diameter that was measured above 100, with the largest diameter being 435, which is why the x axis has values past 400. 

3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
```{r}
missing_data <- vancouver_trees %>%
  summarise(across(everything(), ~sum(is.na(.), na.rm = TRUE)))
print(missing_data)

missing_data_long <- missing_data %>%
pivot_longer(everything(), names_to = "variable", values_to = "missing_count")

ggplot(missing_data_long, aes(x = variable, y = missing_count)) +
geom_bar(stat = "identity", fill = "cornflowerblue", color = "black") + 
  labs(title = "Missing Values per Variable", x = "Variable", y = "Number of Missing Values") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
```

I chose to find the missing values for this dataset because while working with the data, I noticed that there were some columns that had missing data and some information on the trees was not collected. There were 5 variables that contained missing values, with the most missing information being in the date planted variable. I chose to represent this as a bar graph because it is easy to visualize which variabes were not missing any data and what variables contained more missing values. 

4.  Explore the relationship between 2 variables in a plot.
```{r}
summary_data <- vancouver_trees %>%
  group_by(genus_name) %>%
  summarize(avg_height = mean(height_range_id, na.rm = TRUE))

ggplot(summary_data, aes(x = genus_name, y = avg_height)) +
  geom_bar(stat = "identity", fill = "#009E73", color = "black")+
  labs(title = "Average Height of Trees of Each Genus", x = "Genus", y = "Height") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 3))
```

I chose to find the average height of the trees in each genus of trees in the dataset. I represented this as a bar graph to show the height of the trees in relation to the other genera. There are many different kinds of trees in the data set and when trying to find the average height of each species, there were too many on the x axis to be able to see and to understand the data. I also chose to represent the height as the average height for each genus because each genus contains many individual trees which when represented as a histogram or plot, could not be easily interpreted, as there were too many individual points to be able to understand. 

6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
```{r}
ggplot(vancouver_trees, aes(y = height_range_id))  +
  geom_boxplot(fill = "#B19CD9", color = "black") + 
  labs(title = "Boxplot of Tree Heights", x = "Trees ", y = "Height Range ID ") +
  theme_minimal() + 
  scale_x_continuous(breaks = NULL)
  
```

I chose to create a boxplot to represent the height range of the trees in the vancouver trees dataset. The boxplot shows the distribution of the height of the trees for each quartile of the data. The middle line inside the box represents the central tendency of the height of the trees. This line is closer to the lower side of the box, which indicates that there is a skew of tree height being shorter, with the median being approximatey 1.8. This boxplot also shows that there are 2 outliers in this dataset, both of which are taller than the other trees. 
<!----------------------------------------------------------------------------->


# Task 3: Choose research questions 

**(4 points)** So far, you have chosen a dataset and gotten familiar with it through exploring the data. You have also brainstormed one research question that interested you (Task 1.4).  Now it's time to pick 4 research questions that you would like to explore in Milestone 2! Write the 4 questions and any additional comments below. 

<!--- *****START HERE***** --->
1. How does tree species diversity vary across different neighbourhoods of Vancouver?
2. Is there a relationship between the age of the trees and how wide their diameter is?
3. Is there a correlation between the species of tree and how big it is, in terms of height and diameter?
4. Since this data started to be collected, have trends of planting trees susceptible to invasive species(American Elm, Green Ash, White Ash, and Black Ash) declined?

  <!----------------------------->
  
# Overall reproducibility/Cleanliness/Coherence Checklist 
  
## Coherence (0.5 points)
  
  The document should read sensibly from top to bottom, with no major continuity errors. An example of a major continuity error is having a data set listed for Task 3 that is not part of one of the data sets listed in Task 1. 

## Error-free code (3 points) 

For full marks, all code in the document should run without error. 1 point deduction if most code runs without error, and 2 points deduction if more than 50% of the code throws an error. 

## Main README (1 point)

There should be a file named `README.md` at the top level of your repository. Its contents should automatically appear when you visit the repository on GitHub.

Minimum contents of the README file:
  
  -   In a sentence or two, explains what this repository is, so that future-you or someone else stumbling on your repository can be oriented to the repository.
-   In a sentence or two (or more??), briefly explains how to engage with the repository. You can assume the person reading knows the material from STAT 545A. Basically, if a visitor to your repository wants to explore your project, what should they know?
  
  Once you get in the habit of making README files, and seeing more README files in other projects, you'll wonder how you ever got by without them! They are tremendously helpful.

## Output (1 point)

All output is readable, recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

(0.5 point deduction if any of the above criteria are not met. 1 point deduction if most or all of the above criteria are not met.)

Our recommendation: right before submission, delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant. Then, after your final commit and push to Github, CHECK on Github to make sure that everything looks the way you intended!

## Tagged release (0.5 points)

You've tagged a release for Milestone 1. 

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
